Connecting inductive definitions, generic models and a modal multiverse for algebra and combinatorics

### 2022-09-19 slides-brixen2022b.tex
typo

\begin{frame}{Constructive red flags}
  A transitive relation~$<$ on a set~$X$ is \ldots
  \begin{itemize}
    \item \ \\[-1.2em]\mbox{\hil{well-founded$_\infty$} iff
    there is no \bad{\uwave{infinite chain}} $x_0 > x_1 > \cdots$.}
    \item \hil{well-founded'$_\infty$} iff
    there is no \bad{\uwave{bad set}} (inhabited and such that for every member there is
    a smaller member).

@@

  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
  \begin{changemargin}{-2em}{-1.5em}
  \begin{enumerate}
  \small
    \item \justifying
    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
    iff every infinite sequence \emph{everywhere} is good \\
    iff the \emph{generic infinite sequence} is good \\
    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
    sequence \emph{everywhere} is good.}
    \pause

@@

    \item \only<2-6>{A ring element~$x$ is nilpotent \\
    iff \emph{everywhere} it is contained in all prime ideals \\
    iff it is contained in the \emph{generic prime ideal} \\
    iff the \emph{theory of prime ideals} proves~$x \in \ppp$.}
    \only<7->{\mbox{A ring element is nilpotent iff all prime
    ideals \emph{everywhere} contain it.}}

    \pause
    \pause
    \item Given a set~$X$, \emph{proximally} there is a surjection~$\NN
    \twoheadrightarrow X$. \\
    \hil{NB:} $(\possible\varphi) \Rightarrow \varphi$, if~$\varphi$ is a
    geometric implication. \\
    \phantom{\hil{NB:}} $(\xpossible\varphi) \Rightarrow \varphi$, if~$\varphi$
    is first-order. \\
    \phantom{\hil{NB:}} $\varphi \Rightarrow (\necessary\varphi)$,\, if~$\varphi$
    is a geometric formula.

@@

    \pause
    \item \emph{Somewhere,} the law of excluded middle holds.
  \end{enumerate}
  \end{changemargin}
\end{frame}

@@

  \mbox{Unrolling this proof gives a program~$(\textsf{Good} \,|\, []) \to
  \prod_{x \? X} \textsf{Acc}(x)$.}
  \vfill
\end{frame}}

\addtocounter{framenumber}{-1}
\begin{frame}{Extracting constructive content}
  \begin{block}{}
    \justifying
    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
    so is~$X \times Y$.
  \end{block}
  \pause

@@

@@

Mit LEM hat jede Sequenz α ein Minimum:
Induktion über einen Wert, der vorkommt (zum Beispiel α(0)).

Mit LEM gilt: ∃n. α(n) < x  oder  nicht.
Im ersten Fall fertig nach IV.
Im zweiten Fall:
  Sei n beliebig. Weiß ¬(α(n) < x).
  Also weiß: ¬(α(n) ≤ x) und ¬¬(x ≤ α(n)).
  Mit LEM also x ≤ α(n).

Mit DC bekomme Folge:
i₀ so, dass α(i₀) Minimum.
i₁ so, dass α(i₁) Minimum unter α(i₀+1), α(i₀+2), ...

Betrachte Folge (β(iṉ))ṉ.
Mit LEM finde Minimum: β(ik).
Dann β(ik) ≤ β(i(k+1)) und natürlich auch α(ik) ≤ α(i(k+1)).

x < y  heißt:  x ≤ y und ¬(y ≤ x)
Behauptung: ≤ af und transitiv, dann < wf.

Betrachte generische absteigende Folge.
Da gut, habe x[i] ≤ x[j].
Wegen Transitivität auch x[i] > x[j],
also x[j] < x[i], also x[j] ≤ x[i] und ¬(x[i] ≤ x[j]).
Widerspruch.

DC
Krull
Countability
Well
Well-founded

@@

Attribution:
Blass(?)
Alexander
Henry
oE countable

>>>>

\begin{frame}{Constructive red flags}
  \vspace*{-0.5em}
  A transitive relation~$<$ on a set~$X$ is \ldots
  \begin{itemize}
    \item \ \\[-1.2em]\mbox{\hil{well-founded$_\infty$} iff
    there is no \bad{\uwave{infinite chain}} $x_0 > x_1 > \cdots$.}
    \item \hil{well-founded'$_\infty$} iff
    there is no \bad{\uwave{bad set}} (inhabited and such that for every member there is
    a smaller member).

@@

  \newcommand{\seqbad}{\only<3>{infinite descending chain}\only<4->{bad set}}
  \begin{changemargin}{-2em}{-1.5em}
  \begin{enumerate}
  \small
    \item \justifying
    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
    iff every infinite sequence \emph{everywhere} is good \\
    iff the \emph{generic infinite sequence} is good \\
    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
    sequence \emph{everywhere} is good.}
    \pause

@@

    \item \only<2-6>{A ring element~$x$ is nilpotent \\
    iff \emph{everywhere} it is contained in all prime ideals \\
    iff it is contained in the \emph{generic prime ideal} \\
    iff the \emph{theory of prime ideals} proves~``$x \in \ppp$''.}
    \only<7->{\mbox{A ring element is nilpotent iff all prime
    ideals \emph{everywhere} contain it.}}

    \pause
    \pause
    \item Given an inhabited set~$X$, \emph{proximally} there is a surjection~$\NN
    \twoheadrightarrow X$ [J--T]. \\
    \hil{NB:} $(\possible\varphi) \Rightarrow \varphi$, if~$\varphi$ is a
    geometric implication. \\
    \phantom{\hil{NB:}} $(\xpossible\varphi) \Rightarrow \varphi$, if~$\varphi$
    is first-order. \\
    \phantom{\hil{NB:}} $\varphi \Rightarrow (\necessary\varphi)$,\, if~$\varphi$
    is a geometric formula.

@@

    \pause
    \item \emph{Somewhere,} the law of excluded middle holds. [Barr]
  \end{enumerate}
  \end{changemargin}
\end{frame}

@@

  \mbox{Unrolling this proof gives a program~$(\textsf{Good} \,|\, []) \to
  \prod_{x:X} \textsf{Acc}(x)$.}
  \vfill
\end{frame}}

\addtocounter{framenumber}{-1}
\begin{frame}{Extracting constructive content}
  \begin{block}{}
    \justifying
    \textbf{Thm.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
    so is~$X \times Y$.
  \end{block}
  \pause

@@

% \begin{document}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
%{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
\begin{frame}{Outlook}
  I learned the idea to study a modal multiverse of toposes from
  \emph{Alexander Oldenziel}, circa 2016. Foreshadowing results:
  \begin{itemize}
    \scriptsize
    \item[1984] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
    Grothendieck.}
    \item[1987] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
    logic and topoi.}
    \item[2013] Shawn Henry. \emph{Classifying topoi and preservation of higher
    order logic by geometric morphisms.}
  \end{itemize}
  Work by Milly Maietti and Steve Vickers on \emph{arithmetic universes} is
  also closely connected. In progress:

  \begin{itemize}
    \item Develop details and formalize.
    \item Determine the precise list of valid modal principles.
    \item Carry out case studies with program extraction.
    \item Incorporate the right adjoints of geometric morphisms.
  \end{itemize}
\end{frame}}

@@

@@
